# Bookdown Tests

A few tests to make sure **bookdown** is rendering things correctly.

```{r setup, include = FALSE}
cran_link <- function(pkg) {
  paste0("[**", pkg, "**](https://CRAN.R-project.org/package=", pkg, ")")
}
```

Test math macros: $\var(\xx)$, $\alpha_\xx$, $\xx \ind \N(0, \sigma_{\tx{pool}})$.

Test figure:
```{r introtestfig, fig.cap = "Test figure."}
plot(1:25, pch = 1:25)
```
The reference to Figure \@ref(fig:introtestfig) is implemented correctly.  Also, the package link to `r cran_link("mniw")` works correctly.

## Recap of Frequentist Inference

- [ ] Likelihood function.

- [ ] Maximum likelihood estimator.

- [ ] Confidence intervals.

- [ ] Hypothesis testing.

- [ ] Model checking.

## Basics of Bayesian Inference

- Prior and posterior distributions.

- Posterior mean.

- Credible intervals.

- Hypothesis testing.

- Model checking.

## Example 1: Normal with Unknown Mean

- Conjugate prior.

- Improper prior.

- Generalization to EFs.

- MSE comparison to MLE.

## Example 2: Normal with Unknown Variance

- Lack of posterior invariance.

- Asymptotic equivalence.

## Decision Theory

## When Not to Use Bayes: Quantile Estimation

## When to Use Bayes: Hierarchical Modeling

